\documentclass[twoside,12pt]{book}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{mathpazo}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{graphicx}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{tikz}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{booktabs}
\usepackage{geometry}
\usepackage{fancyhdr}
\usepackage{titlesec}

\geometry{
  papersize={6in,9in},
  margin=0.75in,
  top=0.75in,bottom=0.75in,
  inner=0.9in,outer=0.65in,
  headheight=14pt
}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE]{\small\textit{2. Game-Theoretic Techniques}}
\fancyhead[RO]{\small\textit{Randomized Algorithms}}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

\definecolor{codegray}{rgb}{0.45,0.45,0.45}
\lstdefinestyle{cppstyle}{
  basicstyle=\ttfamily\scriptsize,
  keywordstyle=\bfseries,
  commentstyle=\itshape\color{codegray},
  numbers=left,
  numberstyle=\tiny\color{codegray},
  numbersep=8pt,
  frame=single,
  framerule=0.4pt,
  rulecolor=\color{codegray},
  backgroundcolor=\color{gray!5},
  breaklines=true,
  tabsize=4,
  showstringspaces=false,
  language=C++,
  morekeywords={nullptr,size_t,auto,const,constexpr,
    unordered_map,vector,pair,mt19937,bool}
}
\lstset{style=cppstyle}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}
\theoremstyle{remark}
\newtheorem{exercise}[theorem]{Exercise}

\begin{document}

\chapter*{2 Game-Theoretic Techniques}
\addcontentsline{toc}{chapter}{2 Game-Theoretic Techniques}
\markboth{2 Game-Theoretic Techniques}{2 Game-Theoretic Techniques}

In this chapter we study several ideas that are basic to the design and
analysis of randomized algorithms.  All the topics in this chapter share a
game-theoretic viewpoint, which enables us to think of a randomized algorithm
as a probability distribution on deterministic algorithms.  This leads to
Yao's Minimax Principle, which can be used to establish a lower bound on the
performance of a randomized algorithm.

%==========================================================================
\section{Game Tree Evaluation}
\label{sec:gametree}
%==========================================================================

We begin with another simple illustration of linearity of expectation, in the
setting of game tree evaluation.  This example will demonstrate a randomized
algorithm whose expected running time is smaller than that of any
deterministic algorithm.  It will also serve as a vehicle for demonstrating a
standard technique for deriving a lower bound on the running time of any
randomized algorithm for a problem.

A \emph{game tree} is a rooted tree in which internal nodes at even distance
from the root are labeled MIN and internal nodes at odd distance are labeled
MAX.  Associated with each leaf is a real number, which we call its
\emph{value}.  The evaluation of the game tree is the following process.
Each leaf returns the value associated with it.  Each MAX node returns the
largest value returned by its children, and each MIN node returns the smallest
value returned by its children.  Given a tree with values at the leaves, the
evaluation problem is to determine the value returned by the root.

The evaluation of game trees plays a central role in artificial intelligence,
particularly in game-playing programs.  The reader may readily associate the
children of a node with the options available to one of the two players in a
game.  The leaves represent the value of the game for either player.  One
player seeks to maximize this value, while the other tries to minimize it.
At each step, an evaluation algorithm chooses a leaf and reads its value.

We study the number of such steps taken by an algorithm for evaluating a game
tree.  We do not charge the algorithm for any other computation.  We will
limit our discussion to the special case in which the values at the leaves are
bits, 0 or~1.  Thus, each MIN node can be thought of as a Boolean AND
operation and each MAX node as a Boolean OR operation.  This special case is
of interest in its own right, having applications in mechanical theorem
proving.

Let $T_{d,k}$ denote a uniform tree in which the root and every internal node
has~$d$ children and every leaf is at distance~$2k$ from the root.  Thus, any
root-to-leaf path passes through $k$ AND nodes (including the root itself)
and $k$ OR nodes, and there are $d^{2k}$ leaves.  An instance of the
evaluation problem consists of the tree $T_{d,k}$ together with a Boolean
value for each of the $d^{2k}$ leaves.  Given an algorithm, we study the
maximum number of steps it takes to evaluate any instance of $T_{d,k}$.

An algorithm begins by specifying a leaf whose value is to be read at the
first step.  Thereafter, it specifies such a leaf at each step, based on the
values it has read on previous steps.  In a deterministic algorithm, the
choice of the next leaf to be read is a deterministic function of the values
at the leaves read so far.  For a randomized algorithm, this choice may be
randomized.

\begin{exercise}
Show that for any deterministic evaluation algorithm, there is an instance of
$T_{d,k}$ that forces the algorithm to read the values on all $d^{2k}$ leaves.
\end{exercise}

\subsection{A Randomized Algorithm}

We now give a simple randomized algorithm and study the expected number of
leaves it reads on any instance of $T_{d,k}$.  To simplify our presentation,
we restrict ourselves to the case $d=2$.  Any deterministic algorithm for this
case can be made to read all $2^{2k}=4^k$ leaves on some instance.  Our
randomized algorithm is based on the following simple observation.  Consider
a single AND node with two leaves.  If the node were to return~0, at least
one of the leaves must contain~0.  A deterministic algorithm inspects the
leaves in a fixed order, and an adversary can therefore always ``hide'' the~0
at the second of the two leaves inspected by the algorithm.  Reading the
leaves in a random order foils this strategy.  With probability~$1/2$, the
algorithm chooses the hidden~0 on the first step, so its expected number of
steps is~$3/2$, which is better than the worst case for any deterministic
algorithm.  Similarly, in the case of an OR node, if it were to return a~1,
then a randomized order of examining the leaves will reduce the expected
number of steps to~$3/2$.

\begin{algorithm}[htbp]
\caption{Randomized evaluation of $T_{2,k}$}
\label{alg:gametree}
\begin{algorithmic}[1]
\Procedure{EvalAND}{$v$}
  \State Pick a child $c$ of $v$ uniformly at random
  \State $val \gets$ \Call{EvalOR}{$c$}
  \If{$val = 0$}
    \State \Return 0
  \EndIf
  \State $val' \gets$ \Call{EvalOR}{other child of $v$}
  \State \Return $val'$
\EndProcedure
\Statex
\Procedure{EvalOR}{$v$}
  \State Pick a child $c$ of $v$ uniformly at random
  \State $val \gets$ \Call{EvalAND}{$c$}
  \If{$val = 1$}
    \State \Return 1
  \EndIf
  \State $val' \gets$ \Call{EvalAND}{other child of $v$}
  \State \Return $val'$
\EndProcedure
\end{algorithmic}
\end{algorithm}

The reader may wonder how the randomized algorithm can benefit if the AND node
were to return~1, or if the OR node were to return~0.  If the two children of
these nodes are leaves, then clearly both leaves must be examined.  The point
is that at an internal AND node in a tree returning a~1, examining the two OR
children (and evaluating their sub-trees) in a random order is still
beneficial.  The two OR children of an AND node must also return~1, and this
is the easy case for the OR nodes.  Similarly, at an internal OR node
returning~0, the two AND children must return~0, and this is the easy case
for the AND nodes.

\begin{theorem}
\label{thm:gametree}
Given any instance of $T_{2,k}$, the expected number of steps for the
randomized algorithm is at most $3^k$.  Since $n=4^k$ leaves, the expected
running time is $n^{\log_4 3}\approx n^{0.793}$.
\end{theorem}

\begin{proof}
We argue by induction on $k$.  The basis ($k=1$) is the single-node case:
for an AND (or OR) node with two leaves, the expected cost is at most~$3/2$.

Assume the expected cost of evaluating any instance of $T_{2,k-1}$ is at
most $3^{k-1}$.  Consider first a tree $T$ whose root is an OR node, each of
whose children is the root of a copy of $T_{2,k-1}$.

\emph{Case 1: OR evaluates to~1.}
At least one child returns~1.  With probability~$1/2$ this child is chosen
first, incurring expected cost at most $3^{k-1}$.  With probability~$1/2$
both sub-trees are evaluated, incurring cost at most $2\times 3^{k-1}$.
The expected cost is at most
\[
  \tfrac{1}{2}\cdot 3^{k-1} + \tfrac{1}{2}\cdot 2\cdot 3^{k-1}
  = \tfrac{3}{2}\cdot 3^{k-1}.
\]

\emph{Case 2: OR evaluates to~0.}
Both children must be evaluated, costing at most $2\times 3^{k-1}$.

Now consider the root of $T_{2,k}$, an AND node.

\emph{Case 1: AND evaluates to~1.}
Both sub-trees return~1.  By the discussion above, the expected cost of
evaluating each OR sub-tree that returns~1 is at most
$\frac{3}{2}\cdot 3^{k-1}$.  By linearity of expectation, the total
expected cost is at most $2\times\frac{3}{2}\times 3^{k-1}=3^k$.

\emph{Case 2: AND evaluates to~0.}
At least one sub-tree returns~0.  With probability~$1/2$ the sub-tree
returning~0 is chosen first, costing at most $2\times 3^{k-1}$ (both its OR
children must be evaluated).  With probability~$1/2$, an additional cost of
$\frac{3}{2}\cdot 3^{k-1}$ may be incurred for the sibling returning~1.
The expected cost is at most
\[
  \tfrac{1}{2}\cdot 2\cdot 3^{k-1}
  + \tfrac{1}{2}\cdot\bigl(2\cdot 3^{k-1}+\tfrac{3}{2}\cdot 3^{k-1}\bigr)
  \le 3^k.
\]
In all cases the expected cost is at most $3^k$.
\end{proof}

\subsection{C++ Implementation}

\begin{lstlisting}[
  caption={Game tree node and randomized evaluator.},
  label={lst:gametree}]
#include <vector>
#include <random>
#include <chrono>
#include <algorithm>
#include <iostream>

class RandEng {
public:
  static RandEng& get() {
    static RandEng e; return e;
  }
  bool coin() {
    return std::uniform_real_distribution<>(
             0.0,1.0)(eng_) < 0.5;
  }
private:
  RandEng()
    : eng_(std::chrono::steady_clock::now()
           .time_since_epoch().count()) {}
  std::mt19937 eng_;
};

enum class NodeType { AND, OR };

struct TreeNode {
  NodeType type;
  bool is_leaf;
  int value;            // valid only if is_leaf
  std::vector<TreeNode*> children;

  TreeNode(NodeType t) : type(t), is_leaf(false),
                         value(0) {}
  TreeNode(int v) : type(NodeType::AND), is_leaf(true),
                    value(v) {}
};

// Randomized evaluation of AND-OR tree
// Returns number of leaves inspected
struct EvalResult { bool val; int leaves; };

EvalResult eval_and(TreeNode* v);
EvalResult eval_or(TreeNode* v);

EvalResult eval_and(TreeNode* v) {
  if (v->is_leaf) return {v->value==1, 1};
  // pick child at random
  int n = v->children.size();
  int pick = std::uniform_int_distribution<>(
                 0, n-1)(RandEng::get().eng_);
  auto r = eval_or(v->children[pick]);
  int cost = r.leaves;
  if (!r.val) return {false, cost};
  // must check the other child
  for (int i=0;i<n;i++) {
    if (i==pick) continue;
    auto r2 = eval_or(v->children[i]);
    cost += r2.leaves;
    if (!r2.val) return {false, cost};
  }
  return {true, cost};
}

EvalResult eval_or(TreeNode* v) {
  if (v->is_leaf) return {v->value==1, 1};
  int n = v->children.size();
  int pick = std::uniform_int_distribution<>(
                 0, n-1)(RandEng::get().eng_);
  auto r = eval_and(v->children[pick]);
  int cost = r.leaves;
  if (r.val) return {true, cost};
  for (int i=0;i<n;i++) {
    if (i==pick) continue;
    auto r2 = eval_and(v->children[i]);
    cost += r2.leaves;
    if (r2.val) return {true, cost};
  }
  return {false, cost};
}
\end{lstlisting}

\begin{lstlisting}[
  caption={Build a random instance of $T_{2,k}$ and evaluate.},
  label={lst:gametree_driver}]
// Build complete binary tree of depth 2k
// with random leaf values
TreeNode* build_tree(int k, auto& eng) {
  if (k==0) {
    int v = std::uniform_int_distribution<>(
                0,1)(eng);
    return new TreeNode(v);
  }
  // root is AND at even depth, OR at odd
  TreeNode* node = new TreeNode(NodeType::AND);
  node->children.push_back(build_tree(k-1,eng));
  node->children.push_back(build_tree(k-1,eng));
  return node;
}

int main() {
  std::mt19937 eng(
    std::chrono::steady_clock::now()
      .time_since_epoch().count());

  for (int k=1;k<=8;k++) {
    int n = 1 << (2*k);  // 4^k leaves
    double total=0;
    int trials=10000;
    for (int t=0;t<trials;t++) {
      TreeNode* root = build_tree(k,eng);
      auto r = eval_and(root);
      total += r.leaves;
    }
    double avg = total/trials;
    double bound = std::pow(3,k);
    std::cout << "k=" << k
              << " n=" << n
              << " avg=" << avg
              << " 3^k=" << bound
              << " n^0.793="
              << std::pow(n,0.79248)
              << "\n";
  }
}
\end{lstlisting}

\begin{remark}
The algorithm above is a Las Vegas algorithm: it always produces the correct
answer, and its running time is the random variable.  The expected running
time of $n^{0.793}$ is strictly better than the $\Theta(n)$ worst case of any
deterministic algorithm.
\end{remark}

%==========================================================================
\section{The Minimax Principle}
\label{sec:minimax}
%==========================================================================

The randomized algorithm of the preceding section has an expected running
time of $n^{0.793}$ on any uniform binary AND-OR tree with $n$ leaves.  Can
we establish that no randomized algorithm can have a lower expected running
time?  We are thus seeking a lower bound on the running time of any
randomized algorithm for this problem.  As a first step toward this end, we
introduce a standard technique for proving such lower bounds: the minimax
principle.  Indeed, it is the only known general technique for proving lower
bounds on the running times of randomized algorithms.

\subsection{Game Theory}

Consider the following game.  Roberta and Charles put their hands behind
their backs and make a sign for one of the following: stone (closed fist),
paper (open palm), and scissors (two fingers).  They then simultaneously
display their chosen sign.  The winner is determined by the following rules:
paper beats stone by wrapping it, scissors beats paper by cutting it, and
stone beats scissors by dulling it.  The loser pays \$1 to the winner, and
the outcome is a draw when the two players choose the same sign.

\begin{figure}[htbp]
\centering
\begin{tabular}{l|rrr}
         & Scissors & Paper  & Stone \\ \hline
Scissors &    0     &   1    &  -1   \\
Paper    &   -1     &   0    &   1   \\
Stone    &    1     &  -1    &   0   \\
\end{tabular}
\caption{Payoff matrix for scissors-paper-stone.}
\label{fig:payoff}
\end{figure}

This is an instance of a \emph{two-person zero-sum game}, and the matrix is
called the \emph{payoff matrix}.  It is called a zero-sum game because the
net amount won by Roberta and Charles is always exactly zero.  In general,
any two-person zero-sum game can be represented by an $n\times m$ payoff
matrix~$\mathbf{M}$ with real entries.  The set of possible strategies of the
row player~$R$ is in correspondence with the rows of~$\mathbf{M}$, and
likewise for the strategies of the column player~$C$.  The entry $M_{ij}$ is
the amount paid by~$C$ to~$R$ when~$R$ chooses strategy~$i$ and~$C$ chooses
strategy~$j$.

Naturally, the goal of the row (column) player is to maximize (minimize) the
payoff.  Assume that this is a zero-information game, in that neither player
has any information about the opponent's strategy.  If~$R$ chooses strategy~$i$,
then she is guaranteed a payoff of $\min_j M_{ij}$, regardless of~$C$'s
strategy.  An optimal strategy for~$R$ is an~$i$ that maximizes
$\min_j M_{ij}$.  Let $V_R = \max_i \min_j M_{ij}$ denote the lower bound
on the value of the payoff to~$R$ when she uses an optimal strategy.  An
optimal strategy for~$C$ is a~$j$ that gives the best possible upper bound on
the payoff from~$C$ to~$R$.  A similar argument establishes that~$C$'s
optimal strategy ensures that his payoff to~$R$ is at most
$V_C = \min_j \max_i M_{ij}$.

\begin{exercise}
Show that the following inequality is valid for all payoff matrices:
\[
  \max_i \min_j M_{ij} \;\le\; \min_j \max_i M_{ij}.
\]
\end{exercise}

In general, the inequality in Exercise~2.1 is strict; for example, in
scissors-paper-stone, $V_R=-1$ and $V_C=1$.  When these two quantities are
equal, the game is said to have a \emph{solution} and the value of the game
is $V=V_R=V_C$.  The solution (or the \emph{saddle-point}) is the specific
choice of (optimal) strategies that lead to this payoff.

\subsection{Mixed Strategies and the Minimax Theorem}

What happens when a game has no solution?  Then there is no clear-cut optimal
strategy for any player.  In fact, any knowledge of the opponent's strategy
can be used to improve the payoff, unlike the case of games with
saddle-points.  An interesting way to get around this is to introduce
randomization in the choice of strategies.  So far we have been talking about
deterministic or \emph{pure strategies}, but now we focus on randomized or
\emph{mixed strategies}.  A mixed strategy is a probability distribution on
the set of possible strategies.  The row player picks a vector
$\mathbf{p}=(p_1,\ldots,p_n)$, which is a probability distribution on the
rows of~$\mathbf{M}$, i.e., $p_i$ is the probability that~$R$ will choose
strategy~$i$; similarly, the column player has a vector
$\mathbf{q}=(q_1,\ldots,q_m)$, which is a probability distribution on the
columns of~$\mathbf{M}$.  The payoff is now a random variable, and its
expectation is given by
\[
  \mathbb{E}[\text{payoff}]
  = \mathbf{p}^T \mathbf{M}\, \mathbf{q}
  = \sum_{i=1}^{n}\sum_{j=1}^{m} p_i\, M_{ij}\, q_j.
\]

As before, using $V_R$ to denote the best possible lower bound on the
expected payoff to~$R$ that can be ensured by choosing a strategy~$\mathbf{p}$,
and using $V_C$ to denote the best possible upper bound on the expected
payoff by~$C$ by choosing a strategy~$\mathbf{q}$, we obtain
\[
  V_R = \max_{\mathbf{p}} \min_{\mathbf{q}} \mathbf{p}^T\mathbf{M}\,\mathbf{q},
  \qquad
  V_C = \min_{\mathbf{q}} \max_{\mathbf{p}} \mathbf{p}^T\mathbf{M}\,\mathbf{q}.
\]

\begin{theorem}[von Neumann's Minimax Theorem]
\label{thm:neumann}
For any two-person zero-sum game specified by a matrix~$\mathbf{M}$,
\[
  \max_{\mathbf{p}} \min_{\mathbf{q}} \mathbf{p}^T\mathbf{M}\,\mathbf{q}
  = \min_{\mathbf{q}} \max_{\mathbf{p}} \mathbf{p}^T\mathbf{M}\,\mathbf{q}.
\]
\end{theorem}

In other words, the largest expected payoff that~$R$ can guarantee by choosing
a mixed strategy is equal to the smallest expected payoff that~$C$ can
guarantee using a mixed strategy.  This common expected payoff value, called
the \emph{value of the game}, is denoted by~$V$.  A pair of mixed strategies
$(\mathbf{p},\mathbf{q})$ which respectively maximize the left-hand side and
minimize the right-hand side of the equation in Theorem~\ref{thm:neumann} is
called a saddle-point, and the two distributions are called \emph{optimal mixed
strategies}.

\begin{theorem}[Loomis' Theorem]
\label{thm:loomis}
For any two-person zero-sum game specified by a matrix~$\mathbf{M}$,
\[
  \max_{\mathbf{p}} \min_j \mathbf{p}^T\mathbf{M}\,\mathbf{e}_j
  = \min_{\mathbf{q}} \max_i \mathbf{e}_i^T\mathbf{M}\,\mathbf{q},
\]
where $\mathbf{e}_k$ denotes a unit vector with a~1 in the $k$th position
and~0s elsewhere.
\end{theorem}

\subsection{Yao's Minimax Principle}

We now describe the application of the above game-theoretic results to
proving lower bounds on the performance of randomized algorithms.  The idea
is to view the algorithm designer as the column player~$C$ and the adversary
choosing the input as the row player~$R$.  The columns correspond to the set
of all possible algorithms; the rows correspond to the set of all possible
inputs (of a fixed size).  The payoff from~$C$ to~$R$ is some real-valued
measure of the performance of an algorithm, such as the running time, the
quality of the solution obtained, communication cost, or space.

Consider a problem where the number of distinct inputs of a fixed size is
finite, as is the number of distinct (deterministic, terminating, and always
correct) algorithms for solving that problem.  A pure strategy for~$C$
corresponds to the choice of a deterministic algorithm, while a pure strategy
for~$R$ corresponds to a specific input.  Notice that an optimal pure strategy
for~$C$ corresponds to an optimal deterministic algorithm, and $V_C$ is the
worst-case running time of any deterministic algorithm for the problem, which
we call the \emph{deterministic complexity} of the problem.

Our interest is in the interpretation of the mixed strategies for the
algorithm designer and the adversary.  A mixed strategy for~$C$ is a
probability distribution over the space of (always correct) deterministic
algorithms, so it is a Las Vegas randomized algorithm.  An optimal mixed
strategy for~$C$ is an optimal Las Vegas algorithm.  A mixed strategy for~$R$
is a distribution over the space of all inputs.  Let us define the
\emph{distributional complexity} of the problem at hand as the expected
running time of the best deterministic algorithm for the worst distribution
on the inputs.

\begin{proposition}[Yao's Minimax Principle]
\label{prop:yao}
Let $\Pi$ be a problem with a finite set~$I$ of input instances (of a fixed
size), and a finite set~$\mathcal{A}$ of deterministic algorithms.  For input
$I\in I$ and algorithm $A\in\mathcal{A}$, let $C(I,A)$ denote the running
time of algorithm~$A$ on input~$I$.  For probability distributions~$\mathbf{p}$
over~$I$ and~$\mathbf{q}$ over~$\mathcal{A}$, let $I_{\mathbf{p}}$ denote a
random input chosen according to~$\mathbf{p}$ and $A_{\mathbf{q}}$ denote a
random algorithm chosen according to~$\mathbf{q}$.  Then:
\[
  \min_{A\in\mathcal{A}} \mathbb{E}[C(I_{\mathbf{p}}, A)]
  \;\le\;
  \max_{I\in I} \mathbb{E}[C(I, A_{\mathbf{q}})].
\]
\end{proposition}

In other words, the expected running time of the optimal deterministic
algorithm for an arbitrarily chosen input distribution~$\mathbf{p}$ is a
lower bound on the expected running time of the optimal (Las Vegas)
randomized algorithm for~$\Pi$.  Thus, to prove a lower bound on the
randomized complexity, it suffices to choose any distribution~$\mathbf{p}$ on
the input and prove a lower bound on the expected running time of
deterministic algorithms for that distribution.  The power of this technique
lies in the flexibility in the choice of~$\mathbf{p}$ and, more importantly,
the reduction to a lower bound on deterministic algorithms.  It is important
to remember that the deterministic algorithm ``knows'' the chosen
distribution~$\mathbf{p}$.

\subsection{C++ Implementation: Computing Mixed Strategy Equilibria}

\begin{lstlisting}[
  caption={Solving a $2\times 2$ zero-sum game for optimal mixed strategies.},
  label={lst:minimax}]
#include <iostream>
#include <cmath>
#include <iomanip>

// Solve a 2x2 zero-sum game with payoff matrix M
// Returns (V, p1, q1) where
//   V    = game value
//   p1   = prob row player plays row 0
//   q1   = prob col player plays col 0
struct GameSolution {
  double V, p1, q1;
};

GameSolution solve_2x2(double M[2][2]) {
  // Row player optimal: p*M[:,0] = p*M[:,1]
  // p*M00 + (1-p)*M10 = p*M01 + (1-p)*M11
  double denom = M[0][0]-M[0][1]-M[1][0]+M[1][1];
  double p1 = 0.5;
  if (std::abs(denom) > 1e-12)
    p1 = (M[1][1]-M[1][0]) / denom;
  p1 = std::max(0.0, std::min(1.0, p1));

  double denom2 = M[0][0]-M[1][0]-M[0][1]+M[1][1];
  double q1 = 0.5;
  if (std::abs(denom2) > 1e-12)
    q1 = (M[1][1]-M[0][1]) / denom2;
  q1 = std::max(0.0, std::min(1.0, q1));

  double V = p1*(q1*M[0][0]+(1-q1)*M[0][1])
           + (1-p1)*(q1*M[1][0]+(1-q1)*M[1][1]);
  return {V, p1, q1};
}

int main() {
  // Scissors-Paper-Stone
  double M[2][2] = {{0, 1}, {-1, 0}};
  auto [V,p1,q1] = solve_2x2(M);
  std::cout << std::fixed << std::setprecision(4);
  std::cout << "Modified scissors-paper-stone:\n";
  std::cout << "  V=" << V
            << " p=(" << p1 << "," << 1-p1 << ")"
            << " q=(" << q1 << "," << 1-q1 << ")\n";
}
\end{lstlisting}

\subsection{Lower Bound for Game Tree Evaluation}

We now apply Yao's Minimax Principle to the problem of game tree evaluation.
The lower bound that results only applies to algorithms that terminate in a
finite number of steps on any input and sequence of random choices.

We once again limit our attention to instances of the AND-OR tree $T_{2,k}$.
We consider the tree as a balanced binary tree of NORs of depth~$2k$.

\begin{exercise}
Show that the tree $T_{2,k}$ is equivalent to a balanced binary tree all of
whose leaves are at distance $2k$ from the root, and all of whose internal
nodes compute the NOR function: a node returns the value~1 if both inputs
are~0, and~0 otherwise.
\end{exercise}

In order to prove a lower bound on the expected number of leaves evaluated by
any randomized algorithm, we have to specify a distribution on instances
(values for the leaves), and then prove a lower bound on the expected running
time of any deterministic algorithm on such inputs.

Let $p=(3-\sqrt{5})/2$.  Each leaf of the tree is independently set to~1
with probability~$p$.  Note that if each input to a NOR node is independently
1 with probability~$p$, then the probability that its output is~1 is the
probability that both its inputs are~0, which is
\[
  (1-p)^2 = \frac{(\sqrt{5}-1)^2}{4} = \frac{6-2\sqrt{5}}{4}
  = \frac{3-\sqrt{5}}{2} = p.
\]
Thus the value of every node of the NOR tree is~1 with probability~$p$, and
the value of a node is independent of the values of all the other nodes on
the same level.

\begin{proposition}
Let $T$ be a NOR tree each of whose leaves is independently set to~1 with
probability~$q$ for a fixed value $q\in[0,1]$.  Let $W(T)$ denote the
minimum, over all deterministic algorithms, of the expected number of steps to
evaluate~$T$.  Then, there is a depth-first pruning algorithm whose expected
number of steps to evaluate~$T$ is $W(T)$.
\end{proposition}

Proposition~2.7 tells us that for the purposes of our lower bound, we may
restrict our attention to depth-first pruning algorithms.  For a depth-first
pruning algorithm evaluating this tree, let $W(h)$ be the expected number of
leaves it inspects in determining the value of a node at distance~$h$ from
the leaves.  Clearly
\[
  W(h) = W(h-1) + (1-p)\times W(h-1),
\]
where the first term represents the work done in evaluating one of the
sub-trees, and the second term represents the work done in evaluating the
other sub-tree (which will be necessary if the first sub-tree returns the
value~0, an event occurring with probability~$1-p$).  Letting $h=\log_2 n$
and solving, we get $W(h)\ge n^{0.694}$.

\begin{theorem}
\label{thm:lowerbound}
The expected running time of any randomized algorithm that always evaluates an
instance of $T_{2,k}$ correctly is at least $n^{0.694}$, where $n=2^{2k}$ is
the number of leaves.
\end{theorem}

\begin{remark}
Our lower bound of $n^{0.694}$ is less than the upper bound of $n^{0.793}$
that follows from Theorem~\ref{thm:gametree}.  The gap is due to the choice
of input distribution; a more careful (and considerably harder) analysis with
a non-independent distribution shows that the algorithm of
Section~\ref{sec:gametree} is in fact optimal.
\end{remark}

%==========================================================================
\section{Randomness and Non-uniformity}
\label{sec:nonuniform}
%==========================================================================

A basic issue in the study of randomized algorithms is the extent to which
randomization is necessary for solving a problem.  When is it possible to
remove the randomization in a randomized algorithm?  The answer depends on a
number of aspects of the problem being solved.  The goal of this section is to
show that this question is more subtle than appears at first, and touches on
the issue of uniformity in algorithms.

\subsection{Boolean Circuits}

\begin{definition}
A \emph{Boolean circuit} with $n$ inputs is a directed acyclic graph with the
following properties:
\begin{enumerate}
  \item There are $n$ vertices of in-degree~0; these are called the inputs and
    are labeled $x_1,x_2,\ldots,x_n$.  There is one vertex with out-degree~0;
    this is called the output.
  \item Every vertex $v$ that is not an input or the output is labeled with one
    Boolean function $b(v)\in\{\mathrm{AND},\mathrm{OR},\mathrm{NOT}\}$.  A
    vertex labeled NOT has in-degree~1.
  \item Under an assignment of input values, each vertex $v$ computes the
    Boolean function $b(v)$ of the values on its incoming edges.
  \item The \emph{size} of a circuit is the number of vertices in it.
\end{enumerate}
\end{definition}

A \emph{randomized circuit} is very similar, except that there may be more
than~$n$ vertices of in-degree~0, partitioned into two classes:
\begin{enumerate}
  \item \textbf{Random inputs}, each of which is assigned an independent
    random value from $\{0,1\}$.
  \item The $n$ \textbf{circuit inputs}, labeled $x_1,\ldots,x_n$.
\end{enumerate}

A randomized circuit is said to \emph{compute} a function $f$ of the inputs
$x_1,\ldots,x_n$ if:
\begin{enumerate}
  \item For inputs for which $f(x_1,\ldots,x_n)=0$, the output is~0 regardless
    of the values of random inputs.
  \item If $f(x_1,\ldots,x_n)=1$, the output is~1 with probability at
    least~$1/2$.
\end{enumerate}

\begin{definition}
A sequence $C=C_1,C_2,\ldots$ of circuits is a \emph{circuit family} for $f$
if $C_n$ has $n$ inputs and computes $f_n(x_1,\ldots,x_n)$ at its output for
all $n$-bit inputs.  The family $C$ is \emph{polynomial-sized} if the size
of $C_n$ is bounded above by $p(n)$ for every $n$, where $p(\cdot)$ is a
polynomial.
\end{definition}

\subsection{Adleman's Theorem}

\begin{theorem}[Adleman's Theorem]
\label{thm:adleman}
If a Boolean function has a randomized, polynomial-sized circuit family, then
it has a polynomial-sized circuit family.
\end{theorem}

\begin{proof}
We show how to turn a given randomized polynomial-sized circuit~$C_n$ for
$f_n(x_1,\ldots,x_n)$ using random inputs $r_1,\ldots,r_m$, into a
deterministic polynomial-sized circuit~$D_n$.

Form a matrix~$\mathbf{M}$ with $2^n$ rows, one for each possible input from
$\{0,1\}^n$.  The matrix has $2^m$ columns, one for each of the possible
$m$-tuples from $\{0,1\}^m$ that the $r_i$ can assume.  The entry $M_{jk}$
is~1 if the setting of $r_1,\ldots,r_m$ corresponding to column~$k$ is a
\emph{witness} for the input corresponding to row~$j$; otherwise, the entry
is~0.  Eliminate all rows corresponding to inputs for which $f_n$ evaluates
to~0.

By definition, at least half the entries of every surviving row equal~1.
Therefore, there must be a column with at least half its entries~1; in other
words, there is an assignment of~0s and~1s to the $r_i$ that serves as a
witness to at least half of the possible inputs.  Let this witness be
$r_1^{(1)},\ldots,r_m^{(1)}$.  Build a circuit $T_1$, which is a copy
of~$C_n$ with the random inputs ``hard-wired'' to
$r_1^{(1)},\ldots,r_m^{(1)}$.  Delete the column corresponding to this
witness, and all rows that had~1s in this column.

The remaining matrix still has the property that every row has at least half
its entries equal to~1.  Repeat: pick a second witness
$r_1^{(2)},\ldots,r_m^{(2)}$ that is a witness for at least half the
remaining inputs and build a circuit~$T_2$.  Continuing in this manner, we
delete all rows while building at most~$n$ circuits $T_1,\ldots,T_n$.

Take the OR of the outputs of $T_1,\ldots,T_n$.  This is a (deterministic)
circuit whose size is $O(n)$ times that of the randomized circuit we started
with.
\end{proof}

\begin{corollary}
$\mathrm{RP}\subseteq\mathrm{P/poly}$.
\end{corollary}

\begin{remark}
The technique in Theorem~\ref{thm:adleman} is the first example we have seen
of \emph{derandomization}---where we take a randomized algorithm and remove
the randomness.  However, the deterministic circuit generated by this process
is one that works for a particular value of~$n$.  The circuit it produces for
$n$ inputs may have very little resemblance to the circuit it produces for
$n+1$ inputs, even if the original randomized circuits were similar.  Any
``practical'' algorithm will exhibit the property of \emph{uniformity}, which
is formalized in the literature as follows.
\end{remark}

\begin{definition}
An algorithm~$A$ is said to use \emph{advice}~$a$ if on an input of length~$n$
it is given the string $a(n)$ on a read-only tape.  We say that $A$ decides a
language~$L$ with advice~$a$ if on an input~$x$ it uses the read-only string
$a(|x|)$ to decide the membership of~$x$ in~$L$.
\end{definition}

The class $\mathrm{P/poly}$ consists of all languages~$L$ that have a
non-uniform polynomial-time algorithm~$A$ such that the length of the advice
string $a(n)$ is bounded by a polynomial in~$n$.

\begin{exercise}
Consider any language $L\subseteq\{0,1\}^*$.  We define a Boolean
function~$f$ corresponding to the language~$L$ as follows.  For any positive
integer~$n$, let $f_n$ be the Boolean function such that for any
$x\in\{0,1\}^n$, $f_n(x)$ assumes the value~1 if $x\in L$ and~0 otherwise.
Show that $L\in\mathrm{P/poly}$ if and only if it has a polynomial-sized
circuit family.
\end{exercise}

In summary, the removal of randomness in Theorem~\ref{thm:adleman} only shows
that this can be done in principle; it is not known how to do this in any
uniform or practical way.

%==========================================================================
\section*{Notes}
\addcontentsline{toc}{section}{Notes}
%==========================================================================

The material of Section~\ref{sec:gametree} is based on a paper of
Snir~[381].  Most of the material in Section~\ref{sec:minimax} is covered in
textbooks on game theory.  Some good sources are the books by
Wang~[213], Luce and Raiffa~[287], and von Neumann and
Morgenstern~[411].  Theorem~\ref{thm:neumann} is due to von
Neumann~[408], and Theorem~\ref{thm:loomis} is due to Loomis~[279].  The
application of the minimax theorems to proving lower bounds on randomized
algorithms was pointed out by Yao~[419].  Proposition~\ref{prop:yao} is also
from~[419].

In our lower bound for game tree evaluation, the principle that any
deterministic algorithm may as well determine the value of one sub-tree before
inspecting any leaves of its sibling (used in
Section~\ref{sec:minimax}) is due to Tarsi~[393].  Saks and
Wigderson~[362] refined the lower bound to show that Snir's algorithm is
optimal among all randomized algorithms.

Theorem~\ref{thm:adleman} is due to Adleman~[1]; a version of this theorem
applicable to circuit families with two-sided error is due to Gill~[166].  The
notion of non-uniformity is studied in depth in the paper by Karp and
Lipton~[245].

%==========================================================================
\section*{Problems}
\addcontentsline{toc}{section}{Problems}
%==========================================================================

\begin{exercise}[2.1]
Show that for any deterministic evaluation algorithm, there is an instance of
$T_{d,k}$ that forces the algorithm to read the values on all $d^{2k}$ leaves.
\end{exercise}

\begin{exercise}[2.2]
Generalize the randomized algorithm and analysis of Section~\ref{sec:gametree}
to trees $T_{d,k}$ for $d>2$.
\end{exercise}

\begin{exercise}[2.3]
(Due to R.~Boppana~[362].)  Consider a uniform rooted tree of height~$h$:
every leaf is at distance~$h$ from the root.  The root, as well as any
internal node, has three children.  Each leaf has a Boolean value associated
with it.  Each internal node returns the value returned by the majority of its
children.  The evaluation problem consists of determining the value of the
root; at each step, an algorithm can choose one leaf whose value it wishes to
read.
\begin{enumerate}[label=(\alph*)]
\item Show that for any deterministic algorithm, there is an instance that
  forces it to read all $n=3^h$ leaves.
\item Consider the recursive randomized algorithm that evaluates two sub-trees
  of the root chosen at random.  If the values returned disagree, it proceeds
  to evaluate the third sub-tree.  Show that the expected number of leaves
  read by this algorithm (on any instance) is at most $n^{0.91}$.
\end{enumerate}
\end{exercise}

\begin{exercise}[2.4]
Determine the value $V_R$ of the following $2\times 2$ matrix game and give
optimal mixed strategies for the two players.
\[
  \mathbf{M} = \begin{pmatrix} 3 & 1 \\ 0 & 2 \end{pmatrix}
\]
\end{exercise}

\begin{exercise}[2.5]
(Due to A.~M.~Karp.)  Let $(a_{ij})$ be an $m\times n$ matrix, let the
vector $(p_1,p_2,\ldots,p_m)$ consist of reals in $[0,1]$ such that
$\sum_i p_i=1$, and let $(q_1,q_2,\ldots,q_n)$ consist of reals in $[0,1]$
such that $\sum_j q_j=1$.  Prove:
\[
  \min_j \sum_i p_i a_{ij}
  \;\le\; \sum_{i,j} p_i a_{ij} q_j
  \;\le\; \max_i \sum_j a_{ij} q_j.
\]
\end{exercise}

\begin{exercise}[2.6]
Use Yao's Minimax Principle to prove a lower bound on the expected running
time of any Las Vegas algorithm for sorting $n$ numbers.
\end{exercise}

\begin{exercise}[2.7]
(Due to A.~M.~Karp.)  You are given an array~$A$ containing $n$ numbers in
sorted order.  In one step, an algorithm may specify an integer $i\in[1,n]$,
and is given the value of $A[i]$ in return.  Determine lower and upper bounds
on the expected number of steps taken by a Las Vegas randomized algorithm to
determine whether or not a given key~$k$ is present in the array.
\end{exercise}

\begin{exercise}[2.8]
(Due to A.~M.~Karp.)  In a graph with $n$ vertices, where $n$ is even, a
perfect matching is a set of $n/2$ edges, no two of which meet at a common
vertex.  Consider a randomized algorithm that takes an $n$-vertex graph as
input and correctly determines whether the graph has a perfect matching.  At
each step the algorithm asks a question of the form ``Is there an edge between
vertex~$i$ and vertex~$j$?''  The complexity of the algorithm is defined as
the maximum, over all $n$-vertex graphs~$G$, of the expected number of
questions $C(n)$ asked when the input graph is~$G$.  Prove: $C(n)=O(n^2)$.
\end{exercise}

\end{document}